Modeling Ionic Polymer-Metal Composites\\ with Space-Time Adaptive Multimesh \emph{hp}-FEM

### 2010-10-28 results.tex
Results section almost re-done

We used two types of initial meshes --- a finer mesh shown 
in Fig.~\ref{fig:mesh}~(c) was used for \emph{p}-adaptivity
and a very coarse initial mesh shown in  Fig.~\ref{fig:mesh}~(a) was 
used for $h$-adaptivity and $hp$-adaptivity.

\begin{figure}[!ht]
  \begin{centering}
  \includegraphics[width=.8\columnwidth]{mesh}
  \caption{\label{fig:mesh} Initial coarse mesh (a),
  	half refined mesh (b) and refined mesh (c). The coarse mesh (a)
	and refined mesh (c) were used in the initial calculations, the latter one
	in case of \emph{p}-adaptivity (including HP\_ANISO\_P). The half-refined mesh (b) was
	used later to optimize the \emph{hp}-adaptive refinement solutions.}
  \end{centering}
\end{figure}

@@

\noindent
It can be seen that $hp$-FEM results in the shorter computing 
time and smaller number
of degrees of freedom (DOF) than the low-order FEM. The same holds true
for H\_ISO and HP\_ISO modes. In fact, in case of H\_ISO the relative error did not
converge to the pre-set threshold value of $0.5\%$.
Therefore, the $h$-FEM solutions will be omitted from the further comparisons.
Instead, only $hp$-FEM solutions on the coarse mesh and $p$-FEM solutions
on the fine mesh will be discussed. It must be also noted that the error
converged to or below $0.5\%$ for all $p$-FEM and $hp$-FEM results, therefore
the error graphs will not be presented.

@@

Running the simulation with different adaptivity modes 
and meshes showed that the multi-mesh $hp$-FEM configuration resulted into
the smallest problems, shortest computing times, and better or the same error 
convergence compared to any single-mesh configuration.
This is illustrated in Figs.~\ref{fig:singlemultidof} 
and Fig.~\ref{fig:singlemulticpu}.

\begin{figure}[!ht]
  \begin{centering}
  \includegraphics[width=\columnwidth]{singlemulti_dof}
  \caption{\label{fig:singlemultidof} Number of degrees of freedom (DOF) as a function 
  of physical time for single-mesh and multi-mesh configurations with the HP\_ISO
  and HP\_ANISO adaptivity modes. Note logarithmic scale on the vertical axis.}
%\vspace{-6mm}
  \end{centering}
\end{figure}

@@

%
%\subsection{Quantitative analysis of the refinement modes}
%
%Based on the results in the previous subsection, H\_ANISO, HP\_ANISO,
%and HP\_ANISO\_H from the \emph{hp/h}-adaptivity group and HP\_ANISO\_P
%from \emph{p}-adaptivity group will be compared
%in terms of NDOFs and cumulative CPU time. 
%In all of the cases, the relative 
%error at each time step remained below
%the threshold which was set to $e_{th}=0.5\%$ between the coarse mesh
%and fine mesh solutions, therefore the error-time plot will not be considered.
%
%\begin{figure}[!ht]
%  \begin{centering}
%  \includegraphics[width=\columnwidth]{cpu}
%  \caption{\label{fig:cpu} Comparative cumulative CPU time for different refinement modes.}
%  \end{centering}
%\end{figure}
%
%\begin{figure}[!ht]
%  \begin{centering}
%  \includegraphics[width=\columnwidth]{dof}
%  \caption{\label{fig:dof} Comparative NDOFs at each time step for 
%  different refinement modes.}
%  \end{centering}
%\end{figure}
%
%Fig.~\ref{fig:cpu} shows the cumulative CPU time for different refinement 
%modes at each time step --- the results were recorded on the same computer
%for the CPU times to be comparable.
%We see that HP\_ANISO\_H and HP\_ANISO require the most
%resources. This can be understood from the fact that in case of these 
%refinement modes the adaptivity algorithm has the largest number of 
%candidates (see Section.~\ref{sec:model} and Fig.~\ref{fig:refinements})
%from which a particular refinement can be chosen.
%At the same time, HP\_ANISO\_P is the fastest compared to the other refinement modes.
%This can be attributed to the finer inital mesh.
%
%Fig.~\ref{fig:dof} shows the NDOFs at each time step.
%It can be seen that the HP\_ANISO results in the 
%smallest problem size --- $N_{DOF} \approx 125$. 
%All the other refinement modes result in a 
%problem size of approximately ($N_{DOF} \approx 250$). Therefore
%the HP\_ANISO is the most efficient in terms of memory consumption.
%
%As we have filtered out
%two notable refinement modes for the given problem --- HP\_ANISO\_P because of the fast
%calculation, and HP\_ANISO because of the small problem size. In the following
%subsection, some optimizations to reduce the CPU time of
%HP\_ANISO and reduce the NDOFs of HP\_ANISO\_P will be considered.

\subsection{Optimization of the HP\_ANISO adaptivity modes}

\begin{figure}[!ht]
  \begin{centering}
  \includegraphics[width=\columnwidth]{refined_cpu}
  \caption{\label{fig:refined-cpu} Cumulative CPU time for HP\_ANISO and HP\_ANISO\_P
  with different initial meshes.}
  \end{centering}
\end{figure}

\begin{figure}[!ht]
  \begin{centering}
  \includegraphics[width=\columnwidth]{refined_dof}
  \caption{\label{fig:refined-dof} NDOFs at each time step for
  HP\_ANISO and HP\_ANISO\_P with different meshes.}
  \end{centering}
\end{figure}

When it comes to a large domain or 3D modeling, the problem
size becomes a very important factor. Namely, 3D full scale solutions tend to use
a lot of memory. Therefore, we consider HP\_ANISO the most
suitable refinement mode for the given problem. 
Hence, some ways to optimize the HP\_ANISO model
to improve the CPU time factor without significantly 
compromising the NDOFs will be considered. 
The desirable output would be HP\_ANISO problem size
close to HP\_ANISO\_P CPU time. 
One way to optimize the problem
is to choose more refined initial mesh. Another way would be to change
the refinement frequency during the solving process. Theoretically it 
could result in a fewer adaptivity steps which are rather expensive in terms
of CPU time. Recall that up to this point,
the initial mesh was loaded in the beginning of each time step. 
True, by employing the optimizations, one must know something about the problem
and its solution beforehand. However, it could still be practial when solving a real problem
in a large domain.

The problem size and CPU time with HP\_ANISO
and HP\_ANISO\_P adaptivities on more refined initial mesh (see Fig.~\ref{fig:mesh}~(c))
compared to the coarse initial mesh (Fig.~\ref{fig:mesh}~(a)) and 
HP\_ANISO\_P solution are shown in Fig.~\ref{fig:refined-cpu} and Fig.~\ref{fig:refined-dof}.
By using initially more refined mesh, the problem solving time
can be reduced in case of HP\_ANISO, at the same time, the problem size increases
as the initial mesh is not necessarily the most optimal one.
However, in this situation, HP\_ANISO\_P and HP\_ANISO perform equally well.

\begin{figure}[!ht]
  \begin{centering}
  \includegraphics[width=\columnwidth]{unreffreq_dof}
  \caption{\label{fig:unreffreq-dof} NDOFs at each time step for
  HP\_ANISO and HP\_ANISO\_P with mesh unrefinement at each time step and at over each
  	time step after first $0.5\ s$ physical solution time.}
  \end{centering}
\end{figure}
The next proposed optimization involved changing the unrefinement frequency.
It is known that the concentration gradient $\nabla C$ changes the most in the initial phase
of the calculation, therefore, the unrefinement after each time step was performed
until $t=0.5\ s$ (physical time). After that, the
unrefinement was performed in $\Delta t = 0.10\ s$ interval.
However, this optimization does not appear to result in a stable solution, 
i.e. the problem size 
does not remain steady, but starts to oscillate depending on the unrefinement
frequency. This is shown in Fig.~\ref{fig:unreffreq-dof}.

Therefore varying the unrefinement frequency will
likely not result in desired results in real applications for given system of equation.
At the same time, by varying the initial mesh size, optimal initial mesh could be found
for both HP\_ANISO and HP\_ANISO\_P refinement modes.

\subsection{HP\_ANISO adaptivity with non-uniform boundary conditions}
\begin{figure}[!ht]
  \begin{centering}
  \includegraphics[width=.75\columnwidth]{cphiorders}
  \caption{\label{fig:cphi-orders} Solutions $C$ and $\phi$
  and corresponding polynomial degrees of the elements at
  $t=0.7\ s$. HP\_ANISO refinement mode was used. The height
  in the solution graphs indicates the value.}
  \end{centering}
\end{figure}

Based on the results, cation concentration and voltage was calculated
for different boundary conditions.
For instance, when voltage is applied as follows
\begin{equation}
  \phi_{\Omega_1}=0.5\frac{x}{width_{\Omega_1}}+0.5,
\end{equation}
the concentration gradient $\nabla C$ and the voltage gradient $\nabla \phi$ are no
longer effectively 1D.
The calculated $C$ and $\phi$ in $\Omega$ and corresponding meshes and polynomial
degrees of the elements are shown in Fig.~\ref{fig:cphi-orders}.
HP\_ANISO refinement mode was used. Notice that the solution
is different to the one in Fig.~\ref{fig:cphi} and the adapted mesh and the
polynomial degrees are also more complicated than in Fig.~\ref{fig:poly}.
It must be noted that in case of non uniform boundary conditions which
results in 2D problem, refined initial mesh was more efficient to use.

>>>>

We used two types of initial meshes --- a finer mesh shown 
in Fig.~\ref{fig:mesh}~(b) was used for \emph{p}-adaptivity
and a very coarse initial mesh shown in  Fig.~\ref{fig:mesh}~(a) was 
used for $h$-adaptivity and $hp$-adaptivity.

\begin{figure}[!ht]
  \begin{centering}
  \includegraphics[width=.6\columnwidth]{mesh}
  \caption{\label{fig:mesh} Initial coarse mesh (a)
  	and refined mesh (b). The coarse mesh (a)
	and refined mesh (b) were used in the initial calculations, the latter one
	in case of \emph{p}-adaptivity (including HP\_ANISO\_P).}
  \end{centering}
\end{figure}

@@

\noindent
It can be seen that $hp$-FEM results in a shorter computing 
time and smaller number
of DOF than the low-order FEM. The same holds true
for H\_ISO and HP\_ISO modes. In fact, in case of H\_ISO the relative error did not
converge to the pre-set threshold value of $0.5\%$.
Therefore, the $h$-FEM solutions will be omitted from the further comparisons.
Instead, only $hp$-FEM solutions on the coarse mesh and $p$-FEM solutions
on the fine mesh will be discussed. It must be also noted that the error
converged to or below $0.5\%$ for all $p$-FEM and $hp$-FEM results, therefore
the graphs of the error as a function of physical time will not be presented.

@@

Running the simulation with different adaptivity modes 
and meshes showed that the multi-mesh $hp$-FEM configuration resulted in
the smallest problems, shortest computing times, and better or the same error 
convergence compared to any single-mesh configuration.
This is illustrated in Figs.~\ref{fig:singlemultidof} 
and Fig.~\ref{fig:singlemulticpu}. Notice that the HP\_ISO with the single-mesh
configuratio did not even converge in terms of number of degrees of freedom --- the
adaptivity resulted in an unreasonable problem size and the calculation was
not finished.

\begin{figure}[!ht]
  \begin{centering}
  \includegraphics[width=\columnwidth]{singlemulti_dof}
  \caption{\label{fig:singlemultidof} Number of DOF as a function 
  of physical time for single-mesh and multi-mesh configurations with the HP\_ISO
  and HP\_ANISO adaptivity modes. Note logarithmic scale on the vertical axis.}
%\vspace{-6mm}
  \end{centering}
\end{figure}

@@

HP\_ANISO also results in the fastest
computing time among $hp$-adaptivity group whereas HP\_ANISO\_P
results in the fastest overall computing time. This is due to the fact
that HP\_ANISO\_P calculation is performed on the refined mesh. 
Regardless, the HP\_ANISO adaptivity mode is the most suitable
for the PNP problem due to the small size and relative fastness compared
to the other adaptivity modes. A way to optimize the computing time
of HP\_ANISO will be considered next.

\subsection{Time stpe control of HP\_ANISO adaptivity}

\begin{figure}[!ht]
  \begin{centering}
  \includegraphics[width=\columnwidth]{cphi_relerr}
  \caption{\label{fig:cphirelerr} Relative difference $e_{C}^n$ and $e_{\phi}^n$.}
  \end{centering}
\end{figure}
\begin{figure}[!ht]
  \begin{centering}
  \includegraphics[width=\columnwidth]{timeadapt_dof}
  \caption{\label{fig:timeadapt_dof} Number of DOF
  as a function of physical time for HP\_ANISO with and without
  time step adaptivity. The markers on the graphs indicate the
  time steps.}
  \end{centering}
\end{figure}
\begin{figure}[!ht]
  \begin{centering}
  \includegraphics[width=\columnwidth]{timeadapt_cpu}
  \caption{\label{fig:timeadapt_cpu} Cumulative CPU times
  as a function of physical time for HP\_ANISO with and without
  time step adaptivity. The markers on the graphs indicate the
  time steps.}
  \end{centering}
\end{figure}

To optimize the calculation time of HP\_ANISO, an adaptive time step control was employed.
The classical PID controller was used~\cite{valli2002control,dubcova2010space}.
Since $C$ and $\phi$ change differently in time as was demonstrated in Figs.~\ref{fig:cphi-1}
and~\ref{fig:cphi-2}, the relative changes between the solutions at different
time steps were monitored:
\begin{eqnarray}
  e_C^n & = & \frac{\lVert C^n-C^{n-1}\rVert}{\lVert C^n \rVert},\\
  e_\phi^n & = & \frac{\lVert \phi^n-\phi^{n-1}\rVert}{\lVert \phi^n \rVert}.
\end{eqnarray}
The relative changes to control the time step was calculated as follows:
\begin{equation}
  e^n=\max\left\{ e_C^n,\ e_\phi^n \right\}.
\end{equation}
If $e^n < \delta$ where $\delta>0$ is a defined tolerance, then the time
step for the next iteration is increased smoothly to
\begin{equation}
  \tau^{n+1}=\left( \frac{e^{n-1}}{e^n} \right)^{k_P}\ \left( \frac \delta e^n \right)^{k_l}\
  \left[ \frac{\left( e^{n-1} \right)^2}{e^n e^{n-2}} \right]^{k_D} \tau^n,
\end{equation}
where parameters are from~\cite{valli2002control}:
\begin{equation}
  k_p=0.075,\quad k_l=0.175,\quad k_D=0.01.
\end{equation}
The tolerance $\delta$ was set to $\delta=0.25$ in the current optimization
example. The calculated $e_C^n$ and $e_\phi^n$ are shown in Fig.~\ref{fig:cphirelerr}.
The HP\_ANISO problem size and computing time with and without time step control
are shown in Figs.~\ref{fig:timeadapt_dof} and~\ref{fig:timeadapt_cpu}. The reader
can notice that the computing time was reduced approximately 3 times when the time
step adaptivity was employed.

\subsection{HP\_ANISO adaptivity with physically more realistic boundary conditions}
\begin{figure}[!ht]
  \begin{centering}
  \includegraphics[width=\columnwidth]{cphiorders}
  \caption{\label{fig:cphi-orders} Solutions $C$ and $\phi$
  and corresponding polynomial degrees of the elements at
  $t=0.7\ s$. HP\_ANISO refinement mode was used. The height
  in the solution graphs indicates the value.}
  \end{centering}
\end{figure}

In real physics calculations, the applied voltage on boundary $\partial\Omega_1$
is not constant. This can be, for instance, due to the high resistance of
the electrodes as explained in~\cite{pugal2009}.
To see how the HP\_ANISO adaptivity works for such situations, the
voltage on the boundary was applied as follows:
\begin{equation}
  \phi_{\Omega_1}\left( x \right)=0.5\left[V \right] \frac{x\left[ m \right ]}{\text{width}_{\Omega_1}\left[ m \right]}+0.5\left[ V \right],
\end{equation}
where $\text{width}_{\Omega_1}$ is the width of the boundary. The given boundary is effectively
a linear incrase of the voltage from $\phi_{\Omega_1}\left(x = 0 \right)=0.5\ V$ to
$\phi_{\Omega_1}\left(x=\text{width}_{\Omega_1}\right) = 1.0\ V$.
Now the concentration gradient $\nabla C$ and the voltage gradient $\nabla \phi$ are no
longer effectively in 1D.

The calculated $C$ and $\phi$ in $\Omega$ and corresponding meshes and polynomial
degrees of the elements at $t=0.7\ s$ are shown in Figs.~\ref{fig:cphi-orders}.
Notice that the solution
is different to the one in Fig.~\ref{fig:cphi-1}. The HP\_ANISO
adaptivity algorithm has particularly increased the polynomial degree
and refined the mesh near $\Omega_1$ where a sharp concentration
peak exists (compare to Fig.~\ref{fig:poly}).
At $t=3.0\ s$, the shape of the solution $C$ is similar to the one
in Fig.~\ref{fig:cphi-2} and therefore the polynomial space and mesh gets adapted
accordingly. This example clearly illustrates how 
the solution of PNP with non-uniform
boundary conditions is very dynamic in time, particularly for $C$,
and how the HP\_ANISO time dependent adaptivity 
finds an optimal mesh and polynomial space to adapt to the dynamics
of the problem.
